---
author: "Emanuel Sonder"
date: "16 6 2023"
output: html_document
params:
  tfName:
    label: "tfName"
    value: MAZ
    input: text
  chIPPeakPath:
    label: "chIPPeakPath"
    value: ./tf_peaks/maz_peaks.narrowPeak
    input: text
  atacPeakPath:
    label: "atacPeakPath"
    value: ./tf_peaks/maz_peaks.narrowPeak
    input: text
  lfcDistPath:
    label: "lfcDistPath"
    value: ./lfc_dist/happloinsufficiency.rds
    input: select
    choices: [./lfc_dist/happloinsufficiency.rds, ./lfc_dist/happloinsufficency.rds]
  outDir:
    label: "outDir"
    value: ./outDir
    input: text
  effectStrengths:
    label: "effectStrengths"
    value: 0.0,0.5,1.0,2.0,4.0
  sampleDir:
    label: "sampleDir"
    value: "./atac_bams"
    input: text
  samples:
    label: "samples"
    value: ENCFF495DQP.rds,ENCFF130DND.rds,ENCFF447ZRG.rds,ENCFF966ELR.rds,ENCFF358GWK.rds,ENCFF963YZH.rds
    input: text
  sampleNames:
    label: "sampleNames"
    value: treat1,treat2,treat3,ctrl1,ctrl2,ctrl3
    input: text
  gcBiasesFiles:
    label: "gcBiasesFiles"
    value: NULL,NULL,NULL,NULL,NULL,NULL
    input: text
  design:
    label: "design"
    value: 1,1,1,-1,-1,-1
    input: text
  simCodeDir:
    label: "simCodeDir"
    value: ../atacSignalSimulation.R
    input: text
  newTitle:
    label: "newTitle"
    value: "Fragment Simulations"
    input: text
  probNfGrp1:
    label: "probNfGrp1" 
    value: NULL
    input: text
  probMonoGrp1:
    label: "probMonoGrp1" 
    value: NULL
    input: text
  probDiGrp1:
    label: "probDiGrp1" 
    value: NULL
    input: text
  probTriGrp1:
    label: "probTriGrp1" 
    value: NULL
    input: text
  probNfGrp2:
    label: "probNfGrp1" 
    value: NULL
    input: text
  probMonoGrp2:
    label: "probMonoGrp1" 
    value: NULL
    input: text
  probDiGrp2:
    label: "probDiGrp1" 
    value: NULL
    input: text
  probTriGrp2:
    label: "probTriGrp1" 
    value: NULL
    input: text
  simGCBias:
    label: "simGCBias"
    value: FALSE
  simFLD:
    label: "simFLD"
    value: FALSE
  lfcParadigm:
    label: "lfcParadigm"
    value: "activation"
    input: textq
  gcBiasSettingName:
    label: "gcBiasSettingName"
    value: "same"
    input: text
  FLDSettingName:
    label: "FLDSettingName"
    value: "same"
    input: text
  varyAtacPeaks:
    label: "varyAtacPeaks"
    value: TRUE
title: "`r params$newTitle`"
---

```{r setup, include=FALSE}
library(Repitools)
library(data.table)
library(BSgenome.Hsapiens.UCSC.hg38)
library(GenomicRanges)
library(ggplot2)
library(BiocParallel)
library(parallel)
source(params$simCodeDir)

knitr::opts_chunk$set(message = FALSE, warning=FALSE)
knitr::opts_chunk$set(echo = TRUE)
```

# Load the necessary data

## 1. Load the Peak coordinates and the enrichments over the input

Here the column names of the peaks have to be provided to the simulation function. If not the generic narrow peak columns are assumed (see arguments: `colNamesChIPPeaks`, `colNamesAtacPeaks`). 
```{r, load peaks, eval=TRUE}
chIPPeakDir <- params$chIPPeakPath
atacPeakDir <- params$atacPeakPath

atacPeaks <- fread(atacPeakDir)
head(atacPeaks, 5)

chIPPeaks <- fread(chIPPeakDir)
head(chIPPeaks, 5)
```

## 2. Load the logFC vs Enrichment Distributions

Here the distribution of the logFC & enrichments in the NR3C1 experiment are loaded. The first column of this data.table needs to be enrichment column and the second the log fold change column (with the name, see argument: lfcCol)
```{r, logFC dist, eval=TRUE}
lfcDist <- readRDS(params$lfcDistPath)
```

```{r, exp design, eval=TRUE}
bamDirs <- file.path(params$sampleDir, params$samples)
```

# Simulate 

## Set the params by group 

Here we dont need any params as we dont change gc bias or fld (just empty data.tables)
```{r params group, eval=TRUE}
paramsGroup1 <- data.table(gcBiases=params$gcBiasesFiles[which(params$design==1)], 
                           prob_nf=params$probNfGrp1, 
                           prob_mono=params$probMonoGrp1,
                           prob_di=params$probDiGrp1, 
                           prob_tri=params$probTriGrp1)

paramsGroup2 <- data.table(gcBiases=params$gcBiasesFiles[which(params$design==-1)], 
                           prob_nf=params$probNfGrp2, 
                           prob_mono=params$probMonoGrp2,
                           prob_di=params$probDiGrp2, 
                           prob_tri=params$probTriGrp2)
paramsGroup1$name <- "group1"
paramsGroup2$name <- "group2"
```

## Simulate Settings

regions to simulate:
```{r, load the genome and coords all chr, eval=TRUE}
which <- GRanges(Rle(paste0("chr", 1:22)), IRanges(start=1, width=536870912))
genome <- BSgenome.Hsapiens.UCSC.hg38
```

```{r, error=TRUE, eval=TRUE}
if(length(params$effectStrengths)>1)
{
  BPPARAM=SnowParam(workers=min(4, length(params$effectStrengths)))
} else{
  BPPARAM=SerialParam()
}

simDatasets <- bplapply(params$effectStrengths, function(es,
                                        bamPaths, 
                                        chIPPeakDir,
                                        atacPeakDir,
                                        sampleNames,
                                        gcBiases,
                                        fracSub,
                                        design,
                                        paramsGroup1,
                                        paramsGroup2,
                                        lfcDist,
                                        which,
                                        genome,
                                        simFLD,
                                        simGCBias,
                                        seed,
                                        tfName,
                                        lfcParadigm,
                                        gcBiasSettingName,
                                        FLDSettingName,
                                        simDir,
                                        outDir,
                                        varyAtacPeaks){ 
  
  source(simDir)
  library(data.table)
  library(Repitools)
  library(GenomicRanges)
  library(GenomeInfoDb)
  library(BSgenome.Hsapiens.UCSC.hg38)
  library(GenomicAlignments)
  
  simData <- simAtacData(bamPaths=bamPaths, 
                         chIPPeakDir=chIPPeakDir,
                         atacPeakDir=atacPeakDir,
                         sampleNames=sampleNames,
                         effectStrength=es,
                         gcBiases=gcBiases,
                         fracSub=fracSub,
                         design=design,
                         paramsGroup1=paramsGroup1,
                         paramsGroup2=paramsGroup2,
                         lfcDist=lfcDist,
                         which=which,
                         genome=genome,
                         simGCBias=simGCBias,
                         simFLD=simFLD,
                         varyAtacPeaks=varyAtacPeaks,
                         colNamesChIPPeaks=c("chr","start", "end", 
                                            "name", "score", "strand",
                                            "signalValue", "pValue", "qValue", "peak"),
                         colNamesAtacPeaks=c("chr","start", "end", 
                                            "name", "score", "strand",
                                            "signalValue", "pValue", "qValue", "peak"),
                         enrColChIPName="signalValue",
                         enrColAtacName="signalValue",
                         lfcCol="lfc",
                         annotationStyle="UCSC",
                         maxReadPerPeak=1e6,
                         seed=42)
  
  folderName <- paste(tfName,
                      lfcParadigm,
                      gcBiasSettingName,
                      FLDSettingName, 
                      es, sep="_")
  if(!dir.exists(file.path(outDir,folderName))) dir.create(file.path(outDir,folderName))
  simDataPath <-  file.path(outDir, folderName, "simfrags.rds")
  saveRDS(simData, simDataPath)
  simFrags <- simData[[1]]
  
 # bed file format: chrom, start, end, name
  bedSim <- simFrags[,c("seqnames", "start", "end", "sample", "gc_content", "strand"), with=FALSE]
  bedSim$strand <- "."
  bedSim$score <- 0L
  bedSim$name <- "."
  bedSimMerged <- copy(bedSim)
  bedSimMerged$sample <- "merged"
  colnames(bedSimMerged) <- c("chrom", "chromStart", "chromEnd", "sample", "gc_content", "name", "score", "strand")

  # save merged bed file
  mergedFilePath <- file.path(outDir, folderName, "mergedSamples.bed")
  setorder(bedSimMerged, chrom, chromStart, chromEnd)
  write.table(bedSimMerged[,c("chrom", "chromStart", "chromEnd", "name", "score", "strand"), with=FALSE], 
              mergedFilePath,
              quote=FALSE, col.names=FALSE,
              sep="\t", 
              row.names=FALSE)

  # save .bed file by sample
  bedSim <- split(bedSim, by="sample")
  lapply(names(bedSim),function(sample){
    setorder(bedSim[[sample]], seqnames, start, end)
    singleFilePaths <- file.path(outDir, folderName, paste0(sample, ".bed"))
    write.table(bedSim[[sample]][,c("seqnames", "start", "end", "name", "score", "strand"), with=FALSE], 
                singleFilePaths,
                quote=FALSE, col.names=FALSE,
                sep="\t", 
                row.names=FALSE)})
  
  gc()
  return(null)
  },
  bamPaths=bamDirs, 
  chIPPeakDir=params$chIPPeakPath,
  atacPeakDir=params$atacPeakPath,
  sampleNames=params$sampleNames,
  gcBiases=params$gcBiasesFiles,
  fracSub=1,
  design=params$design,
  paramsGroup1=paramsGroup1,
  paramsGroup2=paramsGroup2,
  lfcDist=lfcDist,
  which=which,
  genome=BSgenome.Hsapiens.UCSC.hg38,
  simFLD=params$simFLD,
  simGCBias=params$simGCBias,
  seed=42,
  tfName=params$tfName,
  lfcParadigm=params$lfcParadigm,
  gcBiasSettingName=params$gcBiasSettingName,
  FLDSettingName=params$FLDSettingName,
  simDir=params$simCodeDir,
  outDir=params$outDir,
  varyAtacPeaks=params$varyAtacPeaks,
  BPPARAM=BPPARAM)
```